---
title: "High-dimensional mediation analysis with fMRI data"
output:
    html_document:
      toc: true
      toc_depth: 2
      number_sections: false
      toc_float:
        collapsed: false
        smooth_scroll: false
      theme: default

date: "<span style='font-size: 13px'> last compiled `r format(Sys.Date(), '%m/%d/%Y')` by Sarah Weinstein (collab w/ Nick Illenberger (NYU), David Smith (Temple), and others)"
---


<style type="text/css">


body{ /* Normal  */
      font-size: 12px;
  }

h1.title {
  font-size: 28px;
  color: Black;
}
h1 { /* Header 1 */
  font-size: 16px;
  color: Black;
}

h2 { /* Header 1 */
  font-size: 15px;
  color: Black;
}

.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# required packages
```{r, include = T, message=F, warning=F}
library(pls)
library(lme4)
library(mediation)
library(RNifti)
library(XML)
library(kableExtra)
```

# Data: ($n=28$ after exclusion criteria)
```{r, include = T}
# data gets formatted in this script:
source("/Users/smweinst/Library/CloudStorage/OneDrive-TempleUniversity/research/collaborations/Smith_Lab/tACS/data/format_voxelwise_single_trial_data.R")
subj = dat$subj_run$subj; length(subj); length(unique(subj))
M = scale(as.matrix(dat$M), center = T, scale = T); print(dim(M))
X = dat$X; print(length(X)); table(X)
Y = dat$Y; print(length(Y)); summary(Y)

# exposure w/ moderator version (separate stimulation + task)
## treat one as treatment, other as moderator
X1_task = ifelse(startsWith(as.character(dat$X), "reward"), 1, 0)
X2_stim = ifelse(endsWith(as.character(dat$X), "rtpj"), 1, 0)
```

# Apply partial least squares (PLS) to fMRI data

- Getting 'directions of variation' (essentially )

```{r, include = T}
pls_M = plsr(M ~ X1_task*X2_stim,
             scale = T, center = T)
M_scores = pls_M$Yscores
colnames(M_scores) = paste0("PLS_M", 1:ncol(M_scores))
rownames(pls_M$Yloadings) = gsub("M.", "",rownames(pls_M$Yloadings))

# format data for input into mediation analysis below
mediation_dat = data.frame(subj, X1_task, X2_stim, M_scores, Y)
out = list()
mediation_summary = list()
moderation_test = list()

mediation_model_dat = as.data.frame(cbind(subj, X_trt = X1_task, X_mdr = X2_stim, M_scores, Y))

head(mediation_model_dat)
```

# Define function for testing moderation effects
- need to do this because the test.modmed() function built into the mediation() package doesn't work for lmer-based mediation models. However, inference should still be valid using the same approach (tweaked code below to make this possible)
```{r, include = T}
# FUNCTION TO TEST MODERATION EFFECTS THAT WORKS FOR LMER (mediation::test.modmed() doesn't work for lmer mediation objects, but the adaptation below does work. not sure if still valid but i don't think it's making assumptions that wouldn't work for mixed effects?)
## edited source code for test.modmed() (from https://github.com/kosukeimai/mediation/blob/master/R/medtests.R)
test_modmed_lmer = function(object, covariates.1, covariates.2, conf.level = 0.95){
  cl <- getCall(object) # extract call (i.e., what goes into mediate())
  
  seed <- .Random.seed
  
  # modify call by replacing covariates with covariates.1 (i.e. when moderator takes on value covariates.1)
  cl$covariates <- covariates.1 # "covariates" argument in mediate() specifies values of subset of covariates in model.m and model.y
  out.1 <- eval(cl, envir = parent.frame()) # re-fit mediate() model. by setting covariates argument to covariates.1, function returns estimates conditional on those covariate values
  
  # same as above but with covariates.2 (i.e., when moderator takes on value covariates.2)
  .Random.seed <- seed
  cl$covariates <- covariates.2
  out.2 <- eval(cl, envir = parent.frame()) # re-fit mediate() model after setting covariates argument to covariates.2
  
  # ACME:
  # difference in point estimates for the average causal mediation effects under covariates.1 vs. covariates.2
  d1.diff <- out.1$d1 - out.2$d1 # d1 values correspond to average causal mediation effect
  
  # difference between simulation draws of ACMEs
  d1.diff.sims <- out.1$d1.sims - out.2$d1.sims
  
  # same thing for average direct effects
  z0.diff <- out.1$z0 - out.2$z0 # difference between ADEs (under covariates.1 vs. covariates.2)
  z0.diff.sims <- out.1$z0.sims - out.2$z0.sims # simulation draws
  
  null.value = 0
  pv <- pval(d1.diff.sims, d1.diff)
  ci <- quantile(d1.diff.sims, c((1 - conf.level)/2, (1 + conf.level)/2))
  names(d1.diff) <- names(null.value) <- "ACME(covariates.1) - ACME(covariates.2)"
  attr(ci, "conf.level") <- conf.level
  res.d1 <- list(statistic = d1.diff, p.value = pv, conf.int = ci,
                 null.value = null.value, alternative = "two.sided",
                 method = "Test of ACME(covariates.1) - ACME(covariates.2) = 0",
                 data.name = paste("estimates from", deparse(substitute(object))))
  
  pv <- pval(z0.diff.sims, z0.diff)
  ci <- quantile(z0.diff.sims, c((1 - conf.level)/2, (1 + conf.level)/2))
  names(z0.diff) <- names(null.value) <- "ADE(covariates.1) - ADE(covariates.2)"
  attr(ci, "conf.level") <- conf.level
  res.z0 <- list(statistic = z0.diff, p.value = pv, conf.int = ci,
                 null.value = null.value, alternative = "two.sided",
                 method = "Test of ADE(covariates.1) - ADE(covariates.2) = 0",
                 data.name = paste("estimates from", deparse(substitute(object))))
  
  class(res.d1) <- class(res.z0) <- "htest"
  res <- list(res.d1, res.z0)
  
  names(res) = c("ACME_test_modmed", "ADE_test_modmed")
  
  return(res)
}
pval <- function(x, xhat){
  ## Compute p-values
  if (xhat == 0) out <- 1
  else {
    out <- 2 * min(sum(x > 0), sum(x < 0)) / length(x)
  }
  return(min(out, 1))
} # from mediation package (https://github.com/kosukeimai/mediation/blob/master/R/medtests.R)
```

# mediation analysis by PLS score
- in the loop below, $j$ iterates over the columns of the scores, and each separate mediation model considers each score
```{r, include = T}
nsim=1000
set.seed(10)
for (j in 1:ncol(M_scores)){ # separate mediation for each feature
  # model mediator
  mod_M = lmer(formula = paste0("PLS_M", j, "~ X_trt*X_mdr + (1|subj)"), dat = mediation_model_dat)
  
  # model outcome
  mod_Y = lmer(formula = paste0("Y ~ PLS_M",j, "*X_mdr + X_trt + (1|subj)"), dat = mediation_model_dat)
  
  # mediation analysis:
  mod_mediation = mediate(model.m = mod_M,
                          model.y = mod_Y,
                          treat = "X_trt",
                          mediator = paste0("PLS_M", j),
                          data = mediation_model_dat, sims = nsim)
  
  # summary of mediation output
  out[[paste0("PLS.M",j)]]$mediation_summary = summary(mod_mediation)
  
  # test for moderation by X_mdr (function defined above)
  out[[paste0("PLS.M",j)]]$moderation_test = test_modmed_lmer(mod_mediation, 
                                                              covariates.1 = list(X_mdr = 1),
                                                              covariates.2 = list(X_mdr = 0))
}
```

# plot results 
```{r, include = T, fig.width = 7, fig.height = 5}
unmod_col = rgb(0.7,0,0,0.5)
mod_col = rgb(0,0,0.6,0.5)
par(mfrow=c(1,1), mar = c(2,2,2,2), oma = c(2,2,2,2))

plot_ACME = function(out, mod = F){
  efct = "ACME"
    efct_var = "d1" # this corresponds to ACME in mediation output
  
  plot(0, type = 'n', xlim = c(1,3.1), ylim = c(-40,40), xaxt = 'n',ylab = "", xlab = "", cex.axis = 0.8, las = 2)
  abline(h=0,lty=1, lwd = 2,col = "lightgrey")
  axis(1, at = 1:length(out), labels = paste0("PLS component # ", 1:length(out)), cex.axis = 0.75)
  
  # ACME (not moderation effect):
  if (mod == F){ 
    mtext(efct, line = 0.5, cex = 1, at = 1, adj = 0, font = 2)
    est = sapply(1:ncol(M_scores), FUN = function(j){
      out[[paste0("PLS.M",j)]]$mediation_summary[[efct_var]]
    })
    CI = sapply(1:ncol(M_scores), FUN = function(j){
      out[[paste0("PLS.M",j)]]$mediation_summary[[paste0(efct_var,".ci")]]
    })
    est_CI = rbind(est, CI)
    
    for (j in 1:length(out)){
      # unmoderated:
      points(x = j, y = est_CI[1,j], pch = 16, col = unmod_col)
      segments(x0 = j, x1 = j, y0 = est_CI[2,j], est_CI[3,j], col = unmod_col)
      segments(x0 = j-0.025, x1 = j+0.025, y0 = est_CI[2,j], est_CI[2,j], col = unmod_col)
      segments(x0 = j-0.025, x1 = j+0.025, y0 = est_CI[3,j], est_CI[3,j], col = unmod_col)
      
    }
    
    colnames(est_CI) = paste0("PLS.M", 1:length(out))
    kable(t(est_CI), row.names = T, col.names = c("ACME", "lower", "upper")) %>%
      add_header_above(c(" " = 2, "95% CI" = 2)) %>%
      kable_styling("condensed", full_width = T)  
    
  }else{ # moderation effect
    mtext(paste0(efct, "(rtpj) - ", efct, "(dlpfc)"), line = 0.5, cex = 1, at = 1, adj = 0, font = 2)
    
    # moderated:
    mod_est = sapply(1:ncol(M_scores), FUN = function(j){
      out[[paste0("PLS.M",j)]]$moderation_test[[paste0(efct,"_test_modmed")]]$statistic
    })
    
    mod_CI = sapply(1:ncol(M_scores), FUN = function(j){
      out[[paste0("PLS.M",j)]]$moderation_test[[paste0(efct,"_test_modmed")]]$conf.int
    })
    
    mod_est_CI = rbind(mod_est, mod_CI)
    
    for (j in 1:length(out)){
      points(x = j, y = mod_est_CI[1,j], pch = 16, col = mod_col)
      segments(x0 = j, x1 = j, y0 = mod_est_CI[2,j], mod_est_CI[3,j], lty = 1, col = mod_col)
      segments(x0 = j-0.025, x1 = j+0.025, y0 = mod_est_CI[2,j], mod_est_CI[2,j], lty = 1, col = mod_col)
      segments(x0 = j-0.025, x1 = j+0.025, y0 = mod_est_CI[3,j], mod_est_CI[3,j], lty = 1, col = mod_col)
    }
    
    colnames(mod_est_CI) = paste0("PLS.M", 1:length(out))
    kable(t(mod_est_CI), row.names = T, col.names = c("ACME(rtpj) - ACME(dlpfc)", "lower", "upper")) %>%
      add_header_above(c(" " = 2, "95% CI" = 2)) %>%
      kable_styling("condensed", full_width = T)  
      
  }

}

plot_ACME(out, mod = F)
plot_ACME(out, mod = T)
```

# indicate values of loading weights within each region
```{r, include = T, fig.width = 10, fig.height = 10}
# create dataframe with label names
cort_label_names_df = as.data.frame(do.call("rbind",strsplit(x = dat$label_names$cort, ": "))); colnames(cort_label_names_df) = c("label","name")
sub_label_names_df = as.data.frame(do.call("rbind",strsplit(x = dat$label_names$sub, ": "))); colnames(sub_label_names_df) = c("label","name")
cort_label_names_df = cort_label_names_df[-which(cort_label_names_df$label==0),]
sub_label_names_df = sub_label_names_df[-which(sub_label_names_df$label==0),]

label_names = list(cort = cort_label_names_df,
                   sub = sub_label_names_df)

par(mfrow=c(1,2), mar = c(1,6,2,1), oma = c(0,10,0,0))
for (pc in 1:3){
  for (cort_sub in c("cort", "sub")){ # 1st column of dat$M_voxelwise_labels is cortical labels, 2nd column is subcortical labels
    unique_lab = sort(unique(dat$M_voxelwise_labels[,paste0(cort_sub, "_lab")])[-which(is.na(unique(dat$M_voxelwise_labels[,paste0(cort_sub, "_lab")])))])
    label_names_temp = label_names[[cort_sub]]$name[which(label_names[[cort_sub]]$label %in% unique_lab)]
    
    plot(0, xlim = c(-0.4,0.4), ylim = c(0,(length(unique_lab) + 1)), xlab = "", ylab = "", axes = F)
    axis(1, line = -1.5, cex.axis = 0.75)
    axis(2, at = 1:length(unique_lab), las = 2, labels = paste0(label_names_temp, " (V = ", table(dat$M_voxelwise_labels[,paste0(cort_sub, "_lab")]),")"), cex.axis=0.6,lty=0, line = -2)
    segments(x0=0,x1=0,y0=0,y1=length(unique_lab)+1, lwd = 2, col = rgb(0,0,0,0.25), lty=3)
    abline(h = 1:length(unique_lab) + 0.5, col = rgb(0,0,0,0.5))
    
    for (lab in unique_lab){
      points(x = pls_M$Yloadings[which(dat$M_voxelwise_labels[,paste0(cort_sub, "_lab")] == lab),pc],
             y = rep(which(unique_lab==lab), length(which(dat$M_voxelwise_labels[,paste0(cort_sub, "_lab")] == lab))),
             pch = 16, col = rgb(0,0.4,0.4, 0.3), cex = 0.65)
    }
    
    if (cort_sub=="cort"){
      mtext(paste0("PLS component # ", pc, " voxelwise weights"), at = -0.5, cex = 0.9,adj = 0, font = 2)
    }
    
    mtext(ifelse(cort_sub=="cort", "cortical ROIs", "subcortical ROIs"),
          line = -1, at = -0.5, cex = 0.8, font = 3,adj = 0)
    
  }
}
```

<!-- # export loading values as nifti -->
<!-- ```{r, include = F, eval = F} -->
<!-- # define a function to put data for remaining voxels back into an object with same dimension as nifti files -->
<!-- # 15457 = number of voxels out of original 325325 that are not masked out or regions that are excluded -->
<!-- dat15457_to_nifti = function(dat15457){ -->
<!--   dat325325 = mask#vector(mode = "numeric", length = 325325) -->
<!--   dat325325[mask==0] = NA -->
<!--   dat325325[mask==1][voxels_exclude==1] = NA -->
<!--   dat325325[mask==1][voxels_exclude==0] = dat15457 -->
<!--   return(dat325325) -->
<!-- } -->


<!-- for (j in 1:ncol(pls_M$Yloadings)){ -->
<!--   pls_M.j = dat15457_to_nifti(pls_M$Yloadings[,j]) -->
<!--   cat("PLS component # ", j, ":\n") -->
<!--   print(summary(pls_M.j)) -->
<!--   cat("\n") -->
<!--   writeNifti(pls_M.j, file = paste0("/Users/smweinst/Library/CloudStorage/OneDrive-TempleUniversity/research/collaborations/Smith_Lab/tACS/analysis/pls_mediation/voxelwise/n28_plsM_weights_nii/pls_M", j,".nii"), template = mask) -->
<!-- } -->
<!-- ``` -->

<!-- file names: -->
<!-- ```{r, include = T} -->
<!-- list.files(paste0("/Users/smweinst/Library/CloudStorage/OneDrive-TempleUniversity/research/collaborations/Smith_Lab/tACS/analysis/pls_mediation/voxelwise/n28_plsM_weights_nii/")) -->
<!-- ``` -->
